---
title: "Vulture Lab Results Exploration"
author: "Chris B"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F)

#Load Packages
list.of.packages <- c( "googledrive",  "plotly", "googlesheets4", "MetBrewer", "dplyr")

# Check you have them and load them
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, ,repos = "http://cran.us.r-project.org")

lapply(list.of.packages, require, character.only = TRUE)

x <- drive_find(n_max = 30)

# Create the file
dir.create("./data")
```


```{r}
# Import data
labs <- read_sheet(drive_get("https://docs.google.com/spreadsheets/d/1GYcqjqLG2SUKMF_s4f7JMGtgOt6snzEzHtFCsiwFiiQ"), col_types="c")

inds <- read_sheet(drive_get("https://docs.google.com/spreadsheets/d/1EbAqY7yPZFJwbXxVCq8cWfzL_ol1TW5pb9GpJ56z79M"), col_types="c")

inds <- inds[,c( "cod_sp", "species")]
inds <- inds[is.na(inds$cod_sp)==F,]
labs <- left_join(labs, inds, by=c("code_sp"="cod_sp"))
```

To date we have processed biological samples with three different methodologies:

- Lead and Mercury using voltammetry and atomic absorption on feathers (Peru)
- Lead, Mercury and others using a scanning electron microscope on feathers (Costa Rica)
- Blood cell counts using EDTA bloods (Costa Rica)

The results are shown below. Any feedback on these results would be much appreciated. 

## Contamination - Voltammetry and atomic absorbtion (Peru)

We processed samples from 10 individuals in Peru - specifically looking for mercury and lead contamination. Assay details are scant, and we have only a single reading per individual.

The assays were performed by MC Quimicalab, and the equipment looked fairly ancient.

### Results

```{r, warning=FALSE, message=FALSE}

dat <- labs[labs$method=="voltammetry and atomic absorption" & is.na(labs$method)==F,]

dat$result <- as.numeric(dat$result)
#boxplot(dat$result~dat$assay_target, las=1, ylab=dat$units[1],
#        xlab="Contaminant")


fig <- plot_ly(y =dat$result,x=dat$assay_target, color=dat$assay_target, 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=dat$local_identifier) %>% 
  layout(yaxis = list(title = 'mg_kg'))

fig
```

So this analysis looks interesting - no mercury detected on feather samples, however 5 of 10 individuals having lead contamination. Levels of >1.5 mg_kg are associated with acute exposure (see Cardiel 2011 below).

The three vultures with the highest contamination levels (OC6, OC4,and OC2) are no longer returning GPS fixes. This could represent a death of the individual or the tracking device. OC1 and OC10 are still functioning (4th and 5th values). 

Splitting the lead results by species gives us:

```{r, message=F, warning=F}
fig <- plot_ly(y =dat$result[dat$assay_target=="lead"],x=dat$species[dat$assay_target=="lead"], color=dat$species[dat$assay_target=="lead"], 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=dat$local_identifier[dat$assay_target=="lead"]) %>% 
  layout(yaxis = list(title = 'mg_kg'))

fig
```

Which suggests the bulk of the contamination is in Greater-yellow headed vultures (Cathartes melambrotus) rather than king vultures (Sarcoramphus). 

**However** I have not found may other examples of this methodology in the literature. Other details of the assay were not provided, although we have some nice photos: [link](https://docs.google.com/document/d/1UezGQGasLwwy5mSU0N-JRzSgxDtR1MlY/edit?usp=sharing&ouid=101375399114641236572&rtpof=true&sd=true).

Assay detection limit unknown... so zeros perhaps do not have much value? 

**Questions** 

- How much faith can we put in this assay?

- What other information would we need from this lab to have "publishable" results?


Cardiel, Iris E., Mark A. Taggart, and Rafael Mateo. "Using Pb–Al ratios to discriminate between internal and external deposition of Pb in feathers." Ecotoxicology and Environmental Safety 74.4 (2011): 911-917.

## Contamination - Electron miscroscope (Costa Rica)

We processed samples from seven individuals from Costa Rica - all of which were king vultures. Here we have multiple replicates per individual (between three and five).

Method: The machine used was a "Scanning Electron Microscope, MEB, Brand: Hitachi, Model: S-3700N." with the samples processed by the Centro de Investigación en Estructuras Microscópicas (CIEMic) de la Universidad de Costa. 

### Results

We had a bunch of different elements analysed:

```{r}
dat <- labs[labs$method=="Scanning Electron Microscope, MEB, Brand: Hitachi, Model: S-3700N." & is.na(labs$method)==F,]

#length(unique(dat$local_identifier))

dat$result <- as.numeric(dat$result)
#boxplot(dat$result~dat$assay_target, las=1, ylab=dat$units[1],
#        xlab="Contaminant")


fig <- plot_ly(y =dat$result,x=dat$assay_target, color=dat$assay_target, 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=dat$local_identifier) %>% 
  layout(yaxis = list(title = 'mg_kg'))


fig
```

Lets zoom in on the individual specific results for lead and mercury. The points are mean values and the lines standard deviations:

```{r}
res <- dat %>%  group_by(local_identifier, assay_target) %>% summarize(mean=mean(result), sd=sd(result))

tmp <- res[res$assay_target=="Lead_Pb",]

fig1 <- plot_ly(data = tmp, x = ~local_identifier, y = ~mean, type = 'scatter', mode = 'markers',

        name = tmp$assay_target[1],

        error_y = ~list(array = sd,

                        color = '#000000'))%>% 
  layout(yaxis = list(title = 'Percentage composition'))


tmp <- res[res$assay_target=="Mercury_Hg",]

fig2 <- plot_ly(data = tmp, x = ~local_identifier, y = ~mean, type = 'scatter', mode = 'markers',

        name = tmp$assay_target[1],

        error_y = ~list(array = sd,

                        color = '#000000'))%>% 
  layout(yaxis = list(title = 'Percentage composition'))


subplot(fig1, fig2, shareY=T)


```

Attenborough is still alive and well! Interestingly, the patterns look the same between both elements. The value of repeated measures through time here would be clear.

**Questions** 

- Is it possible to calculate the magnitude of contamination using molecular weights? (convert from percentages)

- Are any of the other elements of interest?

- Is it an issue that we are not using the (more complex) total and methyl-mercury assays?


## Blood cell counts

We had blood cell parameters quantified for 5 individuals in Costa Rica. 

### Results

Differentials first:

```{r}
dat <- labs[labs$method=="Cell Count" & is.na(labs$method)==F & labs$units=="%",]
table(dat$units)
#length(unique(dat$local_identifier))

dat$result <- as.numeric(dat$result)
#boxplot(dat$result~dat$assay_target, las=1, ylab=dat$units[1],
#        xlab="Contaminant")
params <- dat$assay_target

fig <- plot_ly(y =dat$result,x=dat$assay_target, color=dat$assay_target, 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=dat$local_identifier) %>% 
  layout(yaxis = list(title = 'mg_kg'))


fig
```

Attenborough has a thrombocyte percentage outside of the range of the other data - this vulture also had high Pb and Hg levels. 

Others:

```{r}
dat <- labs[labs$method=="Cell Count" & is.na(labs$method)==F & labs$units!="%" & !(labs$assay_target %in% params),]

dat$result <- as.numeric(dat$result)
#boxplot(dat$result~dat$assay_target, las=1, ylab=dat$units[1],
#        xlab="Contaminant")
tmp <- dat[dat$assay_target=="Erythrocytes",]
fig1 <- plot_ly(y =tmp$result,x=tmp$assay_target,
                name=paste(tmp$assay_target[1], tmp$units[1]),
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))

tmp <- dat[dat$assay_target=="Hemoglobin",]
fig2 <- plot_ly(y =tmp$result,x=tmp$assay_target,
                name=paste(tmp$assay_target[1], tmp$units[1]), 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))

tmp <- dat[dat$assay_target=="VCM",]
fig3 <- plot_ly(y =tmp$result,x=tmp$assay_target,
                name=paste(tmp$assay_target[1], tmp$units[1]), 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))

tmp <- dat[dat$assay_target=="MCC",]
fig4 <- plot_ly(y =tmp$result,x=tmp$assay_target, 
                name=paste(tmp$assay_target[1], tmp$units[1]),
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))

tmp <- dat[dat$assay_target=="HMC",]
fig5 <- plot_ly(y =tmp$result,x=tmp$assay_target, 
                name=paste(tmp$assay_target[1], tmp$units[1]),
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))


tmp <- dat[dat$assay_target=="leucocytes",]
fig6 <- plot_ly(y =tmp$result,x=tmp$assay_target, 
                name=paste(tmp$assay_target[1], tmp$units[1]),
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))

tmp <- dat[dat$assay_target=="thrombocytes",]
fig7 <- plot_ly(y =tmp$result,x=tmp$assay_target, 
                name=paste(tmp$assay_target[1], tmp$units[1]), 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))
tmp <- dat[dat$assay_target=="protein_total",]
fig8 <- plot_ly(y =tmp$result,x=tmp$assay_target,
                name=paste(tmp$assay_target[1], tmp$units[1]), 
               type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8, hovertext=tmp$local_identifier) %>% 
  layout(yaxis = list(title = tmp$units[1], rangemode="tozero"))


subplot(fig1,fig2,fig3,fig4,
        fig5,fig6,fig7,fig8, nrows=2)

```

Interestingly, Attenborough again also has elevated leucocytes and thrombocytes. The others all seems to be in a reasonable range. Could this be a cost of an exposure event? 


**Question** Are there any specific categories we should keep an eye on?

**Question** Is there a database for other bird species?


